Summer 2025 ToDo

### 2025-09-15 Relations/Paper/Well-foundedness.tex
Update to the well-foundedness section

An abstract reduction $\to_R$ is strongly normalizing
if and only if the converse relation is well-founded.
In this section, we will compare several definitions of well-foundedness, which depend on the properties in Defintion~\ref{def:WFproperties} below.

For $P \subseteq A$, let $\ol{P} = \{ x \in A \mid x \notin P\}$ denote the complement of $P$.
\begin{definition}\label{def:WFproperties}
  Let $R \subseteq A \times A$.  We write $Rxy$ when $(x,y) \in R$.
  \begin{enumerate}
    \item $P \subseteq A$ is \emph{inductive}
    if, for each $x \in A$, $(\forall y. Ryx \to y \in P)$ implies $x \in P$.

    \item $x \in A$ is \emph{accessible} if $x \in P$ for each inductive $P$.

@@

    \item $s : \nat \to A$ is \emph{$R$-decreasing} if for all $k$, $(s(k+1),s(k)) \in R$. 
  \end{enumerate}
\end{definition}
The definition of a $R$-decreasing sequence is repeated from Section~\ref{sec:Definitions}, as it is used extensively in this section.

\begin{definition}
  Let $R \subseteq A \times A$. We write $Rxy$ when $(x,y) \in R$.
  \begin{enumerate}
    \item $R$ is \emph{accessibly well-founded} ($\WFacc$) if every element is accessible:
      \[
        \forall x \in A.\; x \text{ is accessible}.
      \]
    \item $R$ is \emph{inductively well-founded} ($\WFind$) if every inductive $P \subseteq A$ is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is inductive} \Rightarrow \forall x \in A.\; x \in P.
      \]
    \item $R$ is \emph{sequentially well-founded} ($\WFseq$) if every sequence contains an index at which it is not decreasing:
      \[
        \forall s : \nat \to A \exists k : \nat.\, \lnot \, R \,(s\,k)\,(s\,(k+1)).
      \]
    \item  $R$ is \emph{well-founded with regard to minimality} ($\WFmin$) if every nonempty subset has a $P$-minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow
        \exists x \in P.\; \text{$x$ is $P$-minimal}.
      \]
    \item  $R$ is \emph{well-founded with regard to minimality for $\lnot \lnot$-closed predicates} ($\WFminDNE$) if every nonempty $\lnot \lnot$-closed subset contains a $P$-minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \left(\ol{\ol P} \subseteq P\right) \Rightarrow
        \exists x \in P.\; \text{$x$ is $P$-minimal}.
      \]
    \item $R$ is \emph{coreductively well-founded} ($\WFcor$) if every coreductive $P \subseteq A$ is universally true:
      \[
        \forall P \subseteq A.\;(\text{$P$ is coreductive}) \Rightarrow \forall x \in A.\; x \in P.
      \]

  \end{enumerate}
\end{definition}
We have formalized the above definitions of well-foundedness in \texttt{WFDefinitions.agda}.

@@

On the other hand, assuming accessibility is $\lnot\lnot$-closed, the other
hypothesis can now be weakened, to only require existence of minimal elements
for $\lnot\lnot$-closed subsets.  This is a significant restriction, which may
be validated in many cases of interests.  In contrast, the original hypothesis
is generally equivalent to every subset being decidable.
(The file WFCounters.agda shows that assuming the full minimality hypothesis
for the standard strict order on the natural numbers entails decidability
of every subset of $\nat$, clearly incompatible with effective semantics.)

One could also consider proving well-foundedness ``up to $\lnot\lnot$'',
asserting that every element is $\lnot\lnot$-accessible,
every sequence $\lnot\lnot$-contains a non-decreasing index,
for every non-empty subset there $\lnot\lnot$-exists a minimal element, etc. These are formalized in the definition below (and in \texttt{WFWeakDefinitions.agda}).

\begin{definition}
  Let $R \subseteq A \times A$. We write $Rxy$ when $(x,y) \in R$.
  \begin{enumerate}
    \item $R$ is \emph{weakly accessibly well-founded} ($\WFaccm$) if accessibility holds up to double negation:
      \[
        \forall x \in A.\; \lnot \lnot  (x \text{ is accessible}).
      \]
    \item $R$ is \emph{weakly inductively well-founded} ($\WFindm$) if every inductive predicate holds for all elements up to double negation:
      \[
        \forall P \subseteq A.\;(\text{$P$ is inductive}) \Rightarrow \forall x \in A.\; \lnot \lnot (x \in P).
      \]
    \item $R$ is \emph{weakly sequentially well-founded} ($\WFseqm$) if no sequence is $R$-decreasing:
      \[
        \forall s:\nat \to A.\; \lnot \text{($s$ is $R$-decreasing)}.
      \]
    \item  $R$ is \emph{weakly well-founded with regard to a minimal element} ($\WFminm$) if for every nonempty subset there $\lnot \lnot$-exists a $P$-minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \lnot \lnot \;
        (\exists x \in P.\; \text{$x$ is $P$-minimal}).
      \]
    \item  $R$ is \emph{weakly well-founded with regard to a minimal element under double negation elimination} ($\WFminDNEm$) if every nonempty $\lnot \lnot$-closed subset $\lnot \lnot$-contains a $P$-minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \left(\ol{\ol P} \subseteq P\right) \Rightarrow \lnot \lnot \;
        (\exists x \in P.\; \text{$x$ is $P$-minimal}).
      \]      
    \item [need to define wfmincor]
  \end{enumerate} 
\end{definition}

>>>>

An abstract reduction $\to_R$ is strongly normalizing
if and only if the converse relation is well-founded.
In this section, we will compare several definitions of well-foundedness,
which depend on the basic notions given in Definition~\ref{def:WFnotions} below.

For the rest of this section, fix $R \subseteq A \times A$.  We write $Rxy$ for $(x,y) \in R$.

In the context of ARS, the reader should think of the converse relation, 
interpreting $Rxy$ as denoting a reduction step $y \to_R x$.

For $P \subseteq A$, let $\ol{P} = \{ x \in A \mid x \notin P\}$ denote the complement of $P$.
\begin{definition}\label{def:WFnotions} \hfill 
   \begin{enumerate}
    \item $x \in A$ is \emph{accessible} if for all $y \in A$, $Ryx$ implies $y$ is accessible. 
      This is an inductive definition, with the base case obtained at 
     those $x$ satisfying $\lnot Ryx$ for all $y$.
    \item $P \subseteq A$ is \emph{inductive}
    if, for each $x \in A$, $(\forall y. Ryx \to y \in P)$ implies $x \in P$.

@@

    % \item $s : \nat \to A$ is \emph{$R$-decreasing} if for all $k$, $(s(k+1),s(k)) \in R$. 
  \end{enumerate}
\end{definition}

The above notions give rise to several distinct definitions of well-foundedness, given below. 
Recall the definition of an $R$-decreasing sequence from Section~\ref{sec:Definitions}:
this is a function $s : \nat \to A$ satisfying $(s(k+1),s(k)) \in R$ for all $k \in \nat$.

\begin{definition} \label{def:WFproperties} \hfill 
  \begin{enumerate}
    \item $R$ is \emph{accessibly well-founded} ($\WFacc$) if every element is accessible:
      \[
        \forall x \in A.\; x \text{ is accessible}
      \]
    \item $R$ is \emph{inductively well-founded} ($\WFind$) if every inductive predicate is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is inductive} \Rightarrow \forall x \in A.\; x \in P
      \]
    \item $R$ is \emph{coreductively well-founded} ($\WFcor$) if every coreductive predicate is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is coreductive} \Rightarrow \forall x \in A.\; x \in P
      \]
    \item  $R$ is \emph{well-founded minimality-wise} ($\WFmin$) if every nonempty subset contains a minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow
        \exists x \in A.\; \text{$x$ is $P$-minimal}
      \]
    \item  $R$ is \emph{well-founded minimality-wise for $\lnot \lnot$-closed predicates} ($\WFminDNE$) if every nonempty $\lnot \lnot$-closed subset contains a minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \ol{\ol P} \subseteq P \Rightarrow
        \exists x \in A.\; \text{$x$ is $P$-minimal}
      \]
   \item $R$ is \emph{sequentially well-founded} ($\WFseq$) if every sequence contains an index at which it is not decreasing:
      \[
        \forall s : \nat \to A\ \exists k : \nat.\, \lnot \, R \,(s\,k)\,(s\,(k+1))
      \]
  \end{enumerate}
\end{definition}
We have formalized the above definitions of well-foundedness in \texttt{WFDefinitions.agda}.

@@

On the other hand, assuming accessibility is $\lnot\lnot$-closed, the other
hypothesis can now be weakened, to only require existence of minimal elements
for $\lnot\lnot$-closed subsets.  This is a significant restriction, which may
be validated in many cases of interests.  In contrast, the original hypothesis
(postulating minimal elements for ALL predicates)
is generally equivalent to every subset being decidable.
For example, the file \texttt{WFCounters.agda} shows that assuming the full minimality hypothesis
for the standard strict order on the natural numbers entails decidability
of every subset of $\nat$ --- which is clearly incompatible with effective semantics.
\apcomment{Make sure the file exists, recover it if necessary!}

One could also consider proving well-foundedness ``up to $\lnot\lnot$'',
asserting that every element is $\lnot\lnot$-accessible,
every sequence $\lnot\lnot$-contains a non-decreasing index,
for every non-empty subset there $\lnot\lnot$-exists a minimal element, etc.
These are formalized in the definition below, and in the corresponding file
\texttt{WFWeakDefinitions.agda}.

\begin{definition} \label{def:WFweakproperties} \hfill 
  \begin{enumerate}
    \item $R$ is \emph{weakly accessibly well-founded} ($\WFaccm$) if accessibility holds for all elements up to double negation:
      \[
        \forall x \in A.\; \lnot \lnot  (x \text{ is accessible})
      \]
    \item $R$ is \emph{weakly inductively well-founded} ($\WFindm$) if every inductive predicate holds for all elements up to double negation:
      \[
        \forall P \subseteq A.\;\text{$P$ is inductive} \Rightarrow \forall x \in A.\; \lnot \lnot (x \in P)
      \]
    \item $R$ is \emph{weakly coreductively well-founded} ($\WFcorm$) if every coreductive predicate holds for all elements up to double negation:
      \[
        \forall P \subseteq A.\;\text{$P$ is coreductive} \Rightarrow \forall x \in A.\; \lnot\lnot(x \in P)
      \]
    \item  $R$ is \emph{weakly well-founded minimality-wise} ($\WFminm$)
      if for every nonempty subset $P$ there $\lnot \lnot$-exists a minimal element in $P$:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \lnot \lnot \;
        (\exists x \in A.\; \text{$x$ is $P$-minimal})
      \]
    \item  $R$ is \emph{weakly well-founded minimality-wise for $\lnot \lnot$-closed predicates} ($\WFminDNEm$)
      if for every nonempty $\lnot \lnot$-closed subset $P$ there $\lnot \lnot$-exists a minimal element in $P$:
      % \[
      %   \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \left(\ol{\ol P} \subseteq P\right) \Rightarrow \lnot \lnot \;
      % (\exists x \in P.\; \text{$x$ is $P$-minimal}) \]
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \ol{\ol P} \subseteq P\Rightarrow \lnot \lnot \;
        (\exists x \in A.\; \text{$x$ is $P$-minimal})
      \]      
  \item $R$ is \emph{weakly sequentially well-founded} ($\WFseqm$) if no sequence is $R$-decreasing:
    \[
      \forall s:\nat \to A.\; \lnot \text{($s$ is $R$-decreasing)}
    \]
\end{enumerate} 
\end{definition}
